---
title: "APTS Design of Studies - Practical 2 Example Solutions"
subtitle: "September 2021"
author: "Dave Woods (<D.Woods@southampton.ac.uk> ; <http://www.southampton.ac.uk/~davew>) <br> Statistical Sciences Research Institute, University of Southampton"
output: 
#beamer_presentation
pdf_document:
#  html_document:
    theme: null
    highlights: null
bibliography: ../notes/apts_doe.bib
#csl: american-statistical-association.csl
number_sections: true
---

\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bvarepsilon}{\boldsymbol{\varepsilon}}
\newcommand{\by}{\boldsymbol{y}}
\newcommand{\rT}{\mathrm{T}}
\newcommand{\Var}{\operatorname{Var}}


<style>
pre {
  font-size: 15px;
}
</style>

```{r, echo=FALSE}
options(width=80)
library(knitr)
library(xtable)
set.seed(1)
knit_hooks$set(no.main = function(before, options, envir) {
    if (before) par(mar = c(4.1, 4.1, 1.1, 1.1), pty = "s")  # smaller margin on top
})
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

**Computer experiments**  
We start by generating a maximin Latin hypercube, following the run of thumb of having 10 times as many runs as input variables. The function <tt>maximinSA_LHS</tt> requires a starting design for the optimisation; we will use a random Latin hypercube of the correct size.

```{r cedesign}
library(DiceDesign)
start.design <- lhsDesign(80, 8)$design
maximin.d <- maximinSA_LHS(start.design)
colnames(maximin.d$design) <- c("rw", "r", "Tu", "Hu", "Tl", "Hl", "L", "Kw")
maximin.d$design
```

To run the experiment, we need to scale our design to the appropriate ranges (<tt>DiceDesign</tt> finds designs on the scale $[0,1]$) and pass each row of the design to the <tt>borehole</tt> function. 

```{r cescale}
library(apts.doe)
l <- c(.05, 100, 63070, 990, 63.1, 700, 1120, 9855)
u <- c(.15, 50000, 115600, 1110, 116, 820, 1680, 12045)
center <- -l / (u - l)
scalev <- 1 / (u - l)
bhdesign <- scale(maximin.d$design, center = center, scale = scalev)
bhdesign
y <- apply(bhdesign, 1, borehole)
y
```

We use <tt>DiceKriging</tt> and the <tt>km</tt> function to fit a GP model to this data. We use our original design on the interval $[0,1]$ so that the correlation parameters are comparable (they are all on the same scale). 

```{r cegp}
library(DiceKriging)
gp <- km(formula = ~., design = maximin.d$design, response = y, control = list(trace = F))
gp
```

The dominant variable is $r_w$, with a large linear coefficient and a small $\theta$ (correlation parameter).
